\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-09 k-planar.tex
Last changes for SODA submission

@@

These theorems have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, $p$-centred colouring  \citep{micek:personal}, and succinct graph representations \citep{bonamy.gavoille.ea:shorter}, some of which
%, and graph encoding \citep{BGP}. 
which we explore in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of \citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

@@

We now define two concepts used in the theorems in \cref{Introduction}: strong products and treewidth. The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if: 
\begin{compactitem}
\item  $v=w$ and $xy\in E(B)$, or 
\item  $x=y$ and $vw\in E(A)$, or  
\item  $vw\in E(A)$ and $xy\in E(B)$. 
\end{compactitem}
A \emph{tree-decomposition} $\mathcal{T}$ of a graph $G$ consists of a tree $T$ and a collection $\mathcal{T}=(B_x:x\in V(T))$ of subsets of $V(G)$ indexed by the nodes of $K$ such that:
\begin{compactenum}[(i)]
\item for every $vw\in E(G)$, there exists some node $x\in V(T)$ with $v,w\in B_x$; and 
\item for every $v\in V(G)$, the induced subgraph $T[v] := T[\{x: v\in B_x\}]$ is connected.  
\end{compactenum}
The \emph{width} of the tree-decomposition $\mathcal{T}$ is $\max\{|B_x|:x\in V(T)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.

@@

\begin{restatable}{thm}{mmg}
  %\begin{thm}
  \label{modern-major-general}
    Let $G$ be a graph having an $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\PP$ be a $(k,d)$-shortcut system for $G$.  Then $G^\PP$ has a $J$-partition of layered width at most $d\ell(k^3+3k)$ for some graph $J$ with treewidth at most $\binom{k+t}{t}-1$.
  %\end{thm}
\end{restatable}

@@

Again, a direct application of \cref{modern-major-general} using \cref{SurfacePartition}(b) implies \cref{k-planar} with a weaker bound on the layered width. We prove the stronger bound in \cref{sec-k-planar}.

@@

For convenience, it will be helpful to assume that $\mathcal{P}$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\PP$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\mathcal{P}}$ but avoids special cases in some of our proofs.

For a tree $T$ rooted at some node $x_0\in V(T)$, we we say that a node $a\in V(T)$ is a \emph{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \emph{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in v(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \emph{strict} $T$-ancestor of $x$ if $a\neq x$. For a rooted tree $T$ and any node $x\in V(T)$, let  
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
be the maximal subtree of $T$ rooted at $x$.

@@

\begin{figure}[htbp]
  \begin{center}
    \includegraphics{figs/tripoddo}
  \end{center}
  \caption{The sets $Y_x$, $F_x$, and $N_x$ associated with $x\in V(T)$
  and the ancestors $a_1,\ldots,a_{t'}$ of $X$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.}
  \label{fig:generalized-tripod}
\end{figure}

@@

\begin{lem}\label{generalized-tripod}
  Let $G$ be a graph; let $\mathcal{L}:=\langle V_1,\ldots,V_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$ where $H$ has treewidth at most $t$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \lemref{nice-decomposition}.  For each $x\in V(T)$, let $V_x := \bigcup_{y\in V(T_x)} Y_y$, $F_x:=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x:=V_x\cup F_x$.  Then, 
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width (with respect to $\mathcal{L}$) at most $\ell$.
    \item\ylabel{separator} For each $x\in V(T)$, there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} For each $x\in V(T)$, there is a set $\{a_1,\ldots,a_{t'}\}$ of $t'\le t$ strict $T$-ancestors of $x$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{lem}

@@

Property~\yref{y-subsets} follows from the fact that $V_x$ is the union of several sets, one of which is $Y_x$.  Property~\yref{containment-i} follows from the definition of $V_x$ and the fact that $V(T_x)\subseteq V(T_a)$. To show Property~\yref{containment-ii} first note that, by \yref{containment-i} it suffices to consider vertices in $w\in F_x=N_x\setminus V_x$. By definition, every vertex $w\in F_x$ is adjacent, in $G$, to a vertex $v\in V_x$.  By \yref{containment-i}, $v\in V_a$, so $w$ is either in $V_a$ or $w$ satisfies the condition $vw\in E(G)$, $v\in V_a$, and $w\not\in V_a$, so $w\in F_a$.  In either case $w\in N_a=V_a\cup F_a$.  Note that none of \yref{y-subsets}--\yref{containment-ii} depends on \yref{ancestor-edge} (which is important, since \yref{containment-i} is used to establish \yref{ancestor-edge} in the following proof).

@@

We are now ready to prove our main result, which we now restate for convenience:

@@

\begin{proof}
Apply \cref{generalized-tripod} to $G$ and let 
let $\mathcal{L}$, $\mathcal{Y}$, $\mathcal{T}$, $T$, $Y_x$, $V_x$, $F_x$, and $N_x$ be defined as in \cref{generalized-tripod}.

@@

  We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\mathcal{P}$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\mathcal{P}$ that crosses $x$.

@@

\begin{proof}
  For the sake of contradiction, assume otherwise, so there exists some $v\in S_x\setminus V_x$. By \yref{y-subsets}, $Y_x\subseteq V_x$, so $v\not\in Y_x$.  Therefore, $\mathcal{P}$ contains a path $P$ that crosses $x$.  The path $P$ contains a subpath $v_0,v_1,\ldots,v_{r}$ such that $v=v_0$ and $v_r\in Y_x$. Since $v\not\in V_x$ and $v_r\in Y_x\subseteq V_x$, \yref{separator}, implies that $v_i\in F_x$ for some $i\in\{0,\ldots,r-1\}$. Now \yref{ancestor-edge} implies $v_i\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  Therefore, either $v\in Y_a$ or $P$ crosses $a$. But this implies that $a(v)$ is a $T$-ancestor of $a$, which is a strict $T$-ancestor of $x$, contradicting the assumption that $v\in S_x$.
\end{proof}

@@

\begin{proof}
  We will count the number of vertices in $S_x\cap V_i$ by upper-bounding the number of vertices contributed to $S_x\cap V_i$ by each vertex $w\in Y_x$. Refer to \cref{contribute}.
  If $w\in Y_x\cap V_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap V_i$.

@@

\begin{proof}
  Suppose, for the sake of contradiction, that neither $x$ nor $y$ is a $T$-ancestor of the other.  Since $xy\in E(J)$, $G^\mathcal{P}$ contains an edge $vw$ with $v\in S_x$ and $w\in S_y$.  Since $xy\in E(G^{\mathcal{P}})$,  $\mathcal{P}$ contains a $vw$-shortcut $P$.  By \clmref{s-subset}, $v\in V_x$ and $w\in V_y$.  By \yref{containment-i}, if neither $x$ nor $y$ is a $T$-ancestor of the other, then $V_x$ and $V_y$ are disjoint.  By \yref{ancestor-edge} $N_x$ and $V_y$ are also disjoint.  By \yref{separator} $P$ contains an internal vertex $v'\in F_x$.  By \yref{ancestor-edge}, $v'\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  But this implies that $v\in S_{a'}$ for some $T$-ancestor $a'$ of $a$, contradicting the assumption that $v\in S_x$.
\end{proof}

@@

  Consider an arbitrary node $x\in V(T)$ whose ancestors are $x_0,\ldots,x_{r-1},x_r$ where $x_0$ is the root of $T$ and $x_r:=x$.  To avoid triple-subscripts in what follows, we abuse notation slightly by using $V_i$, $F_i$, and $N_i$,  as shorthands for $V_{x_i}$, $F_{x_i}$ and $N_{x_i}$, respectively.

  If $x_\delta\in C_x$, it is because $x_\delta x'\in E(J)$ for some $T$-descendant $x'$ of $x$.  This implies $G^{\mathcal{P}}$ contains an edge $vw$ with $v\in S_{x'}$ and $w\in S_{x_\delta}=S_\delta$.  This implies that $\mathcal{P}$ a $vw$-shortcut $P_{vw}$.  Let $v'$ be the second-last vertex of $P_{vw}$ (so $v'w\in E(G)$).

@@

  Let $w_0,w_1,\ldots,w_p$ be the path that begins $w_0:=v'$ and then follows the subpath of $P_{ww'}$ that begins at $w_1:=w$ and ends at $w_p:=w''$.  For each $i\in\{0,\ldots,p\}$, let $s_i=\max\{j\in\{1,\ldots,r\}: \{w_0,\ldots,w_i\}\subseteq V_{j}\}\}$, and let $a_i=x_{s_i}$.  Note that $s_0,\ldots,s_p$ is a non-increasing sequence and $a_0,\ldots,a_p$ is a sequence of nodes of $T$ whose distance from the root is non-increasing.

@@

  At this point, the proof of \cref{modern-major-general} is almost immediate from \clmref{general-width} and \clmref{general-bag-size}, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain an edge $vw$ with $v\in V_i$ and $w\in V_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle V_0',\ldots,V_h'\rangle$ in which $V_i'=\bigcup_{j=ki}^{ki+k-1} V_i$.  This increases the layered width given by \clmref{general-width} from $d\ell(k^2+3)$ to $d\ell(k^3+3k)$.  Therefore $G$ has an $H$-partition of layered width at most $d\ell(k^3+3k)$ in which $H$ has treewidth at most $\binom{k+t}{t}$, completing the proof of \cref{modern-major-general}.
\end{proof}

@@

\begin{lem}[\citep{dujmovic.joret.ea:planar}]
\label{qn}
If a graph $G$ has an $H$-partition of layered width $\ell$, then 
$$\qn(G)\le 3 \ell \, \qn(H) + \floor{\tfrac{3}{2}\ell}\enspace.$$
\end{lem}

@@

\begin{cor}
\label{pck}
For every $k$-planar graph $G$, 
%for some integer $t\leq \binom{k+4}{3}-1$, 
%$$\chi_p(G)\le (18k^2+30k)(p+1) (t+1)\tbinom{p+t}{t}\enspace.$$
$$\chi_p(G)\le (18k^2+30k)(p+1) \binom{k+4}{3} \binom{p+ \binom{k+4}{3}-1}{ \binom{k+4}{3}-1}\enspace.$$
%\note{DW}{I previously wrote ``for some integer $t\leq \binom{k+4}{3}-1$'' thinking that $\binom{p+t}{t}$ might not be monotone with $t$, but it is because of the $+t$.}
%Every $k$-planar graph has a layered $H$-partition of layered width at most $18k^2 + 30k$ in which $H$ has treewidth at most $\binom{k+4}{3}-1$. %$\treewidth$.
%Every graph $G$ that has a layered $H$-partition of layered width at most $\ell$ in which $H$ has treewidth at most $t$ satisfies $\chi_p(G)\le \ell (p+1) (t+1)\binom{p+t}{t}$. 
\end{cor}

\begin{cor}
\label{pc1}
For every $1$-planar graph $G$, 
$$\chi_p(G)\le 20 (p+3)(p+2)(p+1)^2\enspace.$$
%120 (p+1) \binom{p+3}{3} 
%Every graph $G$ that has a layered $H$-partition of layered width at most $\ell$ in which $H$ has treewidth at most $t$ satisfies $\chi_p(G)\le \ell (p+1) (t+1)\binom{p+t}{t}$. 
%Every 1-planar graph has a layered $H$-partition of layered width at most 30 where $H$ is planar and has treewidth at most 3.
\end{cor}

\begin{cor}
\label{pcgk}
For every $(g,k)$-planar graph $G$,  
%for some integer $t\leq \binom{k+5}{4}-1$, 
%$$\chi_p(G) \le \max\{2g,3\}\cdot(6k^2 + 10k) (p+1) (t+1)\tbinom{p+t}{t}\enspace.$$
$$\chi_p(G) \le \max\{2g,3\}\cdot(6k^2 + 10k) (p+1) \binom{k+5}{4}\binom{p+\binom{k+5}{4}-1}{\binom{k+5}{4}-1}\enspace.$$
%O(gk^2)\times p^{O(k^3)}$. 
%%Every graph $G$ that has a layered $H$-partition of layered width at most $\ell$ in which $H$ has treewidth at most $t$ satisfies $\chi_p(G)\le \ell (p+1) (t+1)\binom{p+t}{t}$. 
%Every $(g,k)$-planar graph has a layered $H$-partition of layered width at most 
%$\max\{2g,3\}\cdot(6k^2 + 10k)$ in which $H$ has treewidth at most $\binom{k+5}{4}-1$.
\end{cor}

@@

\begin{cor}
Every $(g,d)$-map graph has $p$-centred chromatic number  
$$\chi_p(G)\le 105\, d(d-3)\, \max\{2g,3\}\, (p+1) \tbinom{p+14}{14}\enspace,$$
and every plane $d$-map graph has $p$-centred chromatic number  
$$\chi_p(G)\le 210\,d(d-3)\, (p+1) \tbinom{p+9}{9}\enspace.$$
\end{cor}

@@

\begin{cor}
Every $(g,\delta)$-string graph has queue-number at most 
$$3 \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)  2^{\binom{ \delta+5}{4}-1} \enspace.$$
and every plane $(0,\delta)$-map graph  has queue-number at most 
$$ 9 (\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)  2^{\binom{ \delta+4}{3}-1} \enspace.$$
%If a graph $G$ has a layered $H$-partition of layered width $\ell$, then 
%$$\qn(G)\le 3 \ell ( 2^t- 1/2) \enspace.$$
\end{cor}

>>>>

\usepackage[mathlines]{lineno}
% \setlength{\linenumbersep}{2em}
\linenumbers
% \rightlinenumbers
% \linenumbers
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
    {\linenomath\csname old#1\endcsname}%
    {\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

@@

These theorems have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, and $p$-centred colouring  \citep{micek:personal}, 
% and succinct graph representations \citep{bonamy.gavoille.ea:shorter}, some of which
%, and graph encoding \citep{BGP}. 
which we explore in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of \citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

@@

We now define two concepts used in the theorems in \cref{Introduction}: strong products and treewidth. The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if: 
\begin{compactitem}
\item  $v=w$ and $xy\in E(B)$, or 
\item  $x=y$ and $vw\in E(A)$, or  
\item  $vw\in E(A)$ and $xy\in E(B)$. 
\end{compactitem}
A \emph{tree-decomposition} $\mathcal{T}$ of a graph $G$ consists of a tree $T$ and a collection $\mathcal{T}=(B_x:x\in V(T))$ of subsets of $V(G)$ indexed by the nodes of $T$ such that:
\begin{compactenum}[(i)]
\item for every $vw\in E(G)$, there exists some node $x\in V(T)$ with $v,w\in B_x$; and 
\item for every $v\in V(G)$, the induced subgraph $T[v] := T[\{x: v\in B_x\}]$ is connected.  
\end{compactenum}
The \emph{width} of the tree-decomposition $\mathcal{T}$ is $\max\{|B_x|:x\in V(T)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.

@@

\begin{restatable}{thm}{mmg}
  %\begin{thm}
  \label{modern-major-general}
    Let $G$ be a graph having an $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\PP$ be a $(k,d)$-shortcut system for $G$.  Then $G^\PP$ has a $J$-partition of layered width at most $d\ell(k^3+3k)$ for some graph $J$ of treewidth at most $\binom{k+t}{t}-1$.
  %\end{thm}
\end{restatable}

@@

Again, a direct application of \cref{modern-major-general} using \cref{SurfacePartition}(b) implies \cref{g-k-planar} with a weaker bound on the layered width. We prove the stronger bound in \cref{sec-k-planar}.

@@

For convenience, it will be helpful to assume that $\mathcal{P}$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\PP$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\mathcal{P}}$ but eliminates special cases in some of our proofs.

For a tree $T$ rooted at some node $x_0\in V(T)$, we we say that a node $a\in V(T)$ is a \emph{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \emph{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in v(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \emph{strict} $T$-ancestor of $x$ if $a\neq x$. 
The \emph{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  For any node $x\in V(T)$, define
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
to be the maximal subtree of $T$ rooted at $x$.

@@

\begin{figure}[htbp]
  \begin{center}
    \includegraphics{figs/tripoddo}
  \end{center}
  \caption{The sets $Y_x$, $F_x$, and $V_x$ associated with $x\in V(T)$
  and the ancestors $a_1,\ldots,a_{t'}$ of $X$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.}
  \label{fig:generalized-tripod}
\end{figure}

@@

\begin{lem}\label{generalized-tripod}
  Let $G$ be a graph; let $\mathcal{L}:=\langle V_1,\ldots,V_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$ where $H$ has treewidth at most $t$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \lemref{nice-decomposition}.  For each $x\in V(T)$, let $V_x := \bigcup_{y\in V(T_x)} Y_y$, $F_x:=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x:=V_x\cup F_x$.  Then, 
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width at most $\ell$ with respect to $\mathcal{L}$.
    \item\ylabel{separator} For each $x\in V(T)$, there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} For each $x\in V(T)$, there is a set $\{a_1,\ldots,a_{t'}\}$ of $t'\le t$ strict $T$-ancestors of $x$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{lem}

@@

Property~\yref{y-subsets} follows from the fact that $V_x$ is the union of several sets, one of which is $Y_x$.  Property~\yref{containment-i} follows from the definition of $V_x$ and the fact that $V(T_x)\subseteq V(T_a)$. To show Property~\yref{containment-ii} first note that, by \yref{containment-i} it suffices to consider vertices $w\in F_x=N_x\setminus V_x$. By definition, every vertex $w\in F_x$ is adjacent, in $G$, to a vertex $v\in V_x$.  By \yref{containment-i}, $v\in V_a$, so $w$ is either in $V_a$ or $w$ satisfies the condition $vw\in E(G)$, $v\in V_a$, and $w\not\in V_a$, so $w\in F_a$.  In either case $w\in N_a=V_a\cup F_a$.  Note that none of \yref{y-subsets}--\yref{containment-ii} depends on \yref{ancestor-edge} (which is important, since \yref{containment-i} is used to establish \yref{ancestor-edge} in the following proof).

@@

We are now ready to prove our main result, which we restate here convenience:

@@

\begin{proof}
Apply \cref{generalized-tripod} to $G$ and let $\mathcal{L}$, $\mathcal{Y}$, $\mathcal{T}$, $T$, $Y_x$, $V_x$, $F_x$, and $N_x$ be defined as in \cref{generalized-tripod}.

@@

  We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\mathcal{P}$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\PP$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\mathcal{P}$ that crosses $x$.

@@

\begin{proof}
  For the sake of contradiction, assume otherwise, so there exists some $v\in S_x\setminus V_x$. By \yref{y-subsets}, $Y_x\subseteq V_x$, so $v\not\in Y_x$.  Therefore, $\mathcal{P}$ contains a path $P$, with $v\in V(P)$, that crosses $x$.  The path $P$ contains a subpath $v_0,v_1,\ldots,v_{r}$ such that $v=v_0$ and $v_r\in Y_x$. Since $v\not\in V_x$ and $v_r\in Y_x\subseteq V_x$, \yref{separator}, implies that $v_i\in F_x$ for some $i\in\{0,\ldots,r-1\}$. Now \yref{ancestor-edge} implies $v_i\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  Therefore, either $v\in Y_a$ or $P$ crosses $a$. But this implies that $a(v)$ is a $T$-ancestor of $a$, which is a strict $T$-ancestor of $x$, contradicting the assumption that $v\in S_x$.
\end{proof}

@@

\begin{proof}
  We count the number of vertices in $S_x\cap V_i$ by upper-bounding the number of vertices contributed to $S_x\cap V_i$ by each vertex $w\in Y_x$. Refer to \cref{contribute}.
  If $w\in Y_x\cap V_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap V_i$.

@@

\begin{proof}
  Suppose, for the sake of contradiction, that neither $x$ nor $y$ is a $T$-ancestor of the other.  Since $xy\in E(J)$, $G^\mathcal{P}$ contains an edge $vw$ with $v\in S_x$ and $w\in S_y$.  Since $vw\in E(G^{\mathcal{P}})$,  $\mathcal{P}$ contains a $vw$-shortcut $P$.  By \clmref{s-subset}, $v\in V_x$ and $w\in V_y$.  By \yref{containment-i}, if neither $x$ nor $y$ is a $T$-ancestor of the other, then $V_x$ and $V_y$ are disjoint.  By \yref{ancestor-edge} $N_x$ and $V_y$ are also disjoint.  By \yref{separator} $P$ contains an internal vertex $v'\in F_x$.  By \yref{ancestor-edge}, $v'\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  But this implies that $a(v)=a'$ so $v\in S_{a'}$ for some $T$-ancestor $a'$ of $a$, contradicting the assumption that $v\in S_x$.
\end{proof}

@@

  Consider an arbitrary node $x\in V(T)$ where $x_0,\ldots,x_r$ is the path from the root $x_0$ of $T$ to $x_r:=x$.  To avoid triple-subscripts in what follows, we abuse notation slightly by using $V_i$, $F_i$, and $N_i$,  as shorthands for $V_{x_i}$, $F_{x_i}$ and $N_{x_i}$, respectively.

  If $x_\delta\in C_x$, it is because $x_\delta x'\in E(J)$ for some $T$-descendant $x'$ of $x$.  This implies $G^{\mathcal{P}}$ contains an edge $vw$ with $v\in S_{x'}$ and $w\in S_{x_\delta}=S_\delta$.  This implies that $\mathcal{P}$ contains a $vw$-shortcut $P_{vw}$.  Let $v'$ be the second-last vertex of $P_{vw}$ (so $v'w\in E(G)$).

@@

  Let $w_0,w_1,\ldots,w_p$ be the path that begins $w_0:=v'$ and then follows the subpath of $P_{ww'}$ that begins at $w_1:=w$ and ends at $w_p:=w''$.  For each $i\in\{0,\ldots,p\}$, let $s_i=\max\{j\in\{1,\ldots,r\}: \{w_0,\ldots,w_i\}\subseteq V_{j}\}\}$, and let $a_i=x_{s_i}$.  Note that $s_0,\ldots,s_p$ is a non-increasing sequence and $a_0,\ldots,a_p$ is a sequence of nodes of $T$ whose distance from the root, $x_0$, of $T$ is non-increasing.

@@

  At this point, the proof of \cref{modern-major-general} is almost immediate from \clmref{general-width} and \clmref{general-bag-size}, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain edges $vw$ with $v\in V_i$ and $w\in V_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle V_0',\ldots,V_h'\rangle$ in which $V_i'=\bigcup_{j=ki}^{ki+k-1} V_i$.  This increases the layered width given by \clmref{general-width} from $d\ell(k^2+3)$ to $d\ell(k^3+3k)$.  Therefore $G$ has an $H$-partition of layered width at most $d\ell(k^3+3k)$ in which $H$ has treewidth at most $\binom{k+t}{t}$, completing the proof of \cref{modern-major-general}.
\end{proof}

@@

\begin{lem}[\citep{dujmovic.joret.ea:planar}]
\label{qn}
If a graph $G$ has an $H$-partition of layered width $\ell$, then 
\[ \qn(G)\le 3 \ell \, \qn(H) + \floor{\tfrac{3}{2}\ell}\enspace.\]
\end{lem}

@@

\begin{cor}
\label{pck}
For every $k$-planar graph $G$, 
%for some integer $t\leq \binom{k+4}{3}-1$, 
%$$\chi_p(G)\le (18k^2+30k)(p+1) (t+1)\tbinom{p+t}{t}\enspace.$$
\[ \chi_p(G)\le (18k^2+30k)(p+1) \binom{k+4}{3} \binom{p+ \binom{k+4}{3}-1}{ \binom{k+4}{3}-1}\enspace. \]
%\note{DW}{I previously wrote ``for some integer $t\leq \binom{k+4}{3}-1$'' thinking that $\binom{p+t}{t}$ might not be monotone with $t$, but it is because of the $+t$.}
%Every $k$-planar graph has a layered $H$-partition of layered width at most $18k^2 + 30k$ in which $H$ has treewidth at most $\binom{k+4}{3}-1$. %$\treewidth$.
%Every graph $G$ that has a layered $H$-partition of layered width at most $\ell$ in which $H$ has treewidth at most $t$ satisfies $\chi_p(G)\le \ell (p+1) (t+1)\binom{p+t}{t}$. 
\end{cor}

\begin{cor}
\label{pc1}
For every $1$-planar graph $G$, 
\[ \chi_p(G)\le 20 (p+3)(p+2)(p+1)^2\enspace. \]
%120 (p+1) \binom{p+3}{3} 
%Every graph $G$ that has a layered $H$-partition of layered width at most $\ell$ in which $H$ has treewidth at most $t$ satisfies $\chi_p(G)\le \ell (p+1) (t+1)\binom{p+t}{t}$. 
%Every 1-planar graph has a layered $H$-partition of layered width at most 30 where $H$ is planar and has treewidth at most 3.
\end{cor}

\begin{cor}
\label{pcgk}
For every $(g,k)$-planar graph $G$,  
%for some integer $t\leq \binom{k+5}{4}-1$, 
%$$\chi_p(G) \le \max\{2g,3\}\cdot(6k^2 + 10k) (p+1) (t+1)\tbinom{p+t}{t}\enspace.$$
\[ \chi_p(G) \le \max\{2g,3\}\cdot(6k^2 + 10k) (p+1) \binom{k+5}{4}\binom{p+\binom{k+5}{4}-1}{\binom{k+5}{4}-1}\enspace. \]
%O(gk^2)\times p^{O(k^3)}$. 
%%Every graph $G$ that has a layered $H$-partition of layered width at most $\ell$ in which $H$ has treewidth at most $t$ satisfies $\chi_p(G)\le \ell (p+1) (t+1)\binom{p+t}{t}$. 
%Every $(g,k)$-planar graph has a layered $H$-partition of layered width at most 
%$\max\{2g,3\}\cdot(6k^2 + 10k)$ in which $H$ has treewidth at most $\binom{k+5}{4}-1$.
\end{cor}

@@

\begin{cor}
Every $(g,d)$-map graph has $p$-centred chromatic number  
\[ \chi_p(G)\le 105\, d(d-3)\, \max\{2g,3\}\, (p+1) \tbinom{p+14}{14}\enspace, \]
and every plane $d$-map graph has $p$-centred chromatic number  
\[ \chi_p(G)\le 210\,d(d-3)\, (p+1) \tbinom{p+9}{9}\enspace. \]
\end{cor}

@@

\begin{cor}
Every $(g,\delta)$-string graph has queue-number at most 
\[3 \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)  2^{\binom{ \delta+5}{4}-1} \enspace.\]
and every plane $(0,\delta)$-map graph  has queue-number at most 
\[ 9 (\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)  2^{\binom{ \delta+4}{3}-1} \enspace.\]
%If a graph $G$ has a layered $H$-partition of layered width $\ell$, then 
%$$\qn(G)\le 3 \ell ( 2^t- 1/2) \enspace.$$
\end{cor}
